---
title: "Long-Term Dormancy"
author: "Stuart E. Jones and Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW
In January 2013, we started an experiment using ~24 strains of bacteria -- some "lab" strains others enviornmental strains -- to estimate survivorship during long-term starvation. Each strain was grown un into logarithmic phase and harvested. The cells were pelleted and washed 5x before being put in replicate 50 mL Falcon tubes. Subsamples were taken on overtime and plated onto R2A plates for enumeration as CFUs. After looking at plots of the data, in became apparent that CFUs were declining at a non-constant rate on semi-log plots. We therefore developed a statistical model to test for whether the the data were better fit by a first-order decay (i.e., log-linear) or a saturating function.  

## 1) SET WORKING DIRECTORY AND LOAD DATA
```{r}
## Retrieve and Set Your Working Directory
rm(list = ls())
getwd()
setwd("~/GitHub/Dimensions/Aim1/DeathCurves/")

## Load Data
obs <- read.csv("data/longtermdormancy_20150806_nocomments.csv", 
                header = TRUE, stringsAsFactors = FALSE)
```

## 2) ESTIMATE COLONY FORMING UNITS (CFU)
```{r}
## Adding 1 to deal with log(0) observations
obs$Abund <- as.numeric(obs$Colonies) * 10 ^ as.numeric(obs$Dilution) + 1
```

## 3) SELECT STRAINS WITH > 10 OBSERVATIONS 
```{r}
strains <- sort(unique(obs$Strain))
strains <- strains[table(obs$Strain)>10]
obs <- obs[obs$Strain%in%strains,]
```

## 4) CREATE MATRIX FOR STORING MODEL OUTPUT
```{r}
summ <- matrix(NA,length(strains)*max(obs$Rep),10)
```

## 5) FIT MODELS
```{r}
pdf('output/decayFits.pdf') # Uncomment to create pdf that will plot data and fits
counter <- 1

for(i in 1:length(strains)){
  strainObs=obs[obs$Strain==strains[i],]
	
	reps=unique(strainObs$Rep)
	for(j in 1:length(reps)){
		
		repObs=strainObs[strainObs$Rep==reps[j],]

		if(nrow(repObs)>3){
			
			start=repObs[1,1]
		
			time=(as.numeric(strptime(repObs$Firstread_date,format="%d-%b-%y",tz="EST"))-
              as.numeric(strptime(start,format="%d-%b-%y",tz="EST")))/(3600*24)
			time2=time*time
	
			# Fit log linear and log quadratic models using glm()	
			logLinCur=glm(log10(repObs$Abund)~time)
			logQuadCur=glm(log10(repObs$Abund)~time+time2)
		
			# pull coefficients, etc. from glm fits
			summ[counter,1]=strains[i]
			summ[counter,2]=reps[j]
			summ[counter,3:4]=coef(logLinCur)[2:1]
			summ[counter,5]=summary(logLinCur)$aic
			summ[counter,6:8]=coef(logQuadCur)[3:1]
			summ[counter,9]=summary(logQuadCur)$aic
		
			# run likelihood ratio test to compare the linear and quadratic model; store p-value in summ
			summ[counter,10]=as.numeric(pchisq(2*(logLik(logQuadCur)-logLik(logLinCur)),df=1,lower.tail=FALSE))
		
			#### add indicator of non-linearity (****) to plot title if quadratic model is better
      
      ### *** Comment/Uncomment following code to make pdf figs*** ###
# 			if(!is.na(summ[counter,10])){
# 				if(as.numeric(summ[counter,10])<0.05){
# 					title=paste(strains[i]," rep ",reps[j],"****")
# 				}else{
# 					title=paste(strains[i]," rep ",reps[j])
# 				}
# 			}else{
# 				title=paste(strains[i]," rep ",reps[j])
# 			}
# 			
# 			plot(time,log10(repObs$Abund),main=title,ylim=c(0,9))
# 			predTime=seq(0,max(time))
# 			lines(predTime,logLinCur$par[2]-logLinCur$par[1]*predTime,lwd=2,lty=2)
# 			lines(predTime,logQuadCur$par[3]-logQuadCur$par[1]*predTime^2-logQuadCur$par[2]*predTime,col='red',lwd=2,lty=2)
      ### *** Comment/Uncomment above code to make pdf figs*** ###		

			counter=counter+1
		}
	}
}

dev.off() # uncomment line and `pdf()` above if you want plots of data and fits 
```

## 6) SUMMARY OF MODEL FITS
```{r}
summ=summ[!is.na(summ[,1]),]
colnames(summ)=c('strain','rep','linearfit_K','linearfit_N0','linearfit_AIC','quadfit_k2','quadfit_k','quadfit_N0','quadfit_AIC','LRT_pvalue')
```

## 7) HOW DID EACH TUBE BEHAVE?
```{r}
#percent of tubes better fit by quadratic (AIC_linear > AIC_quadratic)
sum(as.numeric(summ[,5])>as.numeric(summ[,9]))/nrow(summ)*100  # 82%

# percent of tubes better fit by quadratic using likelihood ratio test
sum(as.numeric(summ[,10])<0.05)/nrow(summ)*100	# 69%

# add bonferroni correction
0.05/nrow(summ)	#  0.00044
sum(as.numeric(summ[,10])<0.0044)/nrow(summ)*100	# 49%

# histogram of likelihood ratio p-values
hist(log10(as.numeric(summ[,10])),breaks=seq(-20,0,1))
abline(v=log10(0.05),lwd=2,lty=2,col='red')
abline(v=log10(0.00044),lwd=2,lty=2,col='green')
legend('topleft',c('alpha=0.05','alpha=0.00044 (bonferroni)'),lty=2,col=c('red','green'),box.lty=0,lwd=2)

# confirm likelihood ratio test is showing what deltaAIC between linear and quadratic would suggest
plot(log10(as.numeric(summ[,10])),as.numeric(summ[,9])-as.numeric(summ[,5]),xlab="log10(LRT pvalue)",ylab="deltaAIC (quadratic - linear)")
abline(v=log10(0.05),lwd=2,lty=2,col='red')
abline(v=log10(0.00044),lwd=2,lty=2,col='green')
```

## 8) IS BEHAVIOR CONSISTENT AMONG REPS (TUBES) WITHIN STRAINS?
```{r}
# look at agreement among reps
repAgreement=matrix(NA,length(strains),4)

repAgreement[,1]=strains
for(i in 1:length(strains)){
  cur=summ[summ[,1]==strains[i],]
	
	repAgreement[i,2]=nrow(cur)
	repAgreement[i,3]=sum(as.numeric(cur[,10])>0.05)
	repAgreement[i,4]=nrow(cur)-sum(as.numeric(cur[,10])>0.05)
}

repAgree=data.frame(strain=repAgreement[,1],Nreps=as.numeric(repAgreement[,2]),Nlinear=as.numeric(repAgreement[,3]),Nquad=as.numeric(repAgreement[,4]),stringsAsFactors=FALSE)
```

## 9) GENERATE TRAITS OUTPUT TABLE
```{r}
# when quadratic is not significantly better by likelihood ratio test report "evolvability" of 0
traitsReport=matrix(NA,nrow(summ),5)
traitsReport[,1]=summ[,1]
traitsReport[,2]=summ[,2]
for(i in 1:nrow(summ)){
  if(as.numeric(summ[i,10])<(0.05/nrow(summ))){
		traitsReport[i,3]=summ[i,7]
		traitsReport[i,4]=summ[i,6]
		traitsReport[i,5]=summ[i,8]
	}else{
		traitsReport[i,3]=summ[i,3]
		traitsReport[i,4]=0
		traitsReport[i,5]=summ[i,4]
	}
}

colnames(traitsReport)=c('strain','rep','decay','evolvability','N0')

# remove the extremely bad fit
traitsReport=traitsReport[-27,]

strains=sort(unique(traitsReport[,1]))
meanDecay=tapply(as.numeric(traitsReport[,3]),traitsReport[,1],FUN=mean)
meanEvolve=tapply(as.numeric(traitsReport[,4]),traitsReport[,1],FUN=mean)
meanN0=tapply(as.numeric(traitsReport[,5]),traitsReport[,1],FUN=mean)

summaryTraitsReport=data.frame(strains=strains,decay=meanDecay,evolvability=meanEvolve,N0=meanN0,stringsAsFactors=FALSE)

write.table(traitsReport,"output/perRepDeathCurveTraits.txt",row.names=FALSE,quote=FALSE)
write.table(summaryTraitsReport,"output/DeathCurveTraits.txt",row.names=FALSE,quote=FALSE)
```
