---
title: "Long-Term Dormancy"
author: "Stuart E. Jones and Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW
In January 2013, we started an experiment using ~24 strains of bacteria -- some "lab" strains others enviornmental strains -- to estimate survivorship during long-term starvation. Each strain was grown un into logarithmic phase and harvested. The cells were pelleted and washed 5x before being put in replicate 50 mL Falcon tubes. Subsamples were taken on overtime and plated onto R2A plates for enumeration as CFUs. After looking at plots of the data, in became apparent that CFUs were declining at a non-constant rate on semi-log plots. Here, we test whether or not the Weibull function can help explain variation in death curves.  

## 1) SET WORKING DIRECTORY, LOAD DATA, LOAD PACKAGE

```{r}
## Retrieve and Set Your Working Directory
rm(list = ls())
getwd()
setwd("~/GitHub/Dimensions/Aim1/")
require("bbmle")
```

##2) LOAD DATA

```{r}
## Load Data
obs <- read.csv("data/longtermdormancy_20170620_nocomments.csv", 
                header = TRUE, stringsAsFactors = FALSE)

## Adding 1 to deal with log(0) observations
obs$Abund <- as.numeric(obs$Colonies) * 10 ^ as.numeric(obs$Dilution) + 1
strains <- sort(unique(obs$Strain))
strains <- strains[table(obs$Strain)>10]
obs <- obs[obs$Strain%in%strains,]

## Time
time <- (as.numeric(strptime(obs$Firstread_date,format="%d-%b-%y",tz = "EST")) - 
        as.numeric(strptime(obs$Dormstart_date,format="%d-%b-%y",tz="EST")))/(3600*24)
time <- ifelse(time == 0, 1, time)
obs <- data.frame(obs, time)

## Adding 1 to deal with log(0) observations
obs$Abund <- as.numeric(obs$Colonies) * 10 ^ as.numeric(obs$Dilution) + 1
strains <- sort(unique(obs$Strain))
strains <- strains[table(obs$Strain)>10]
obs <- obs[obs$Strain%in%strains,]
```

## 3) PICK REPRESENTATIVE STRAINS

```{r}
KBS0701 <- obs[(obs$Strain == "KBS0701"),] # Arthrobacter
KBS0714 <- obs[(obs$Strain == "KBS0714"),] # Micrococcus
KBS0802 <- obs[(obs$Strain == "KBS0802"),] # Pseudomonas
KBS0812 <- obs[(obs$Strain == "KBS0812"),] # Bacillus
```

## 4) FOUR-PANEL PLOT WITHOUT FUNCTIONS

```{r}
png(filename="~/GitHub/Dimensions/Aim1/output/death.wo.fitting.png",
    width = 2400, height = 2400, res = 96*2)

# par(mfrow=c(2,2))
# par(mar = c(5, 7, 5, 7))
# par(oma = c(5,4,0,0))

par(mfrow = c(2,2),
          oma = c(5,4,0,0) + 0.5,
          mar = c(0,3,3,1) + 0.5)

# KBS0701
plot(KBS0701$time, KBS0701$Abund, pch = 22, 
          bg = "white", col = "black", lwd = 6, cex = 6, log = "y", xlab = NA, ylab = NA,
          xlim = c(-100, 1050), ylim = c(1, 100000000), yaxt = "n", xaxt = "n")
box(lwd = 6)

axis(side = 2, lwd.ticks = 6, cex.axis = 3, las = 1,
     labels = expression(10^0, 10^2, 10^4, 10^6, 10^8),
     at = c(10^0, 10^2, 10^4, 10^6, 10^8))
 
axis(side = 4, lwd.ticks = 6, cex.axis = 3, las = 1,
    at=c(10^0, 10^2, 10^4, 10^6, 10^8), labels = F)
 
axis(side = 1, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

axis(side = 3, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

# KBS0714
plot(KBS0714$time, KBS0714$Abund, pch = 22, 
          bg = "white", col = "black", lwd = 6, cex = 6, log = "y", xlab = NA, ylab = NA,
          xlim = c(-100, 1050), ylim = c(1, 1000000000), yaxt = "n", xaxt = "n")
box(lwd = 6)

axis(side = 2, lwd.ticks = 6, cex.axis = 3, las = 1,
     at=c(10^0, 10^2, 10^4, 10^6, 10^8), labels = F)
 
axis(side = 4, lwd.ticks = 6, cex.axis = 3, las = 1,
    at=c(10^0, 10^2, 10^4, 10^6, 10^8), labels = F)
 
axis(side = 1, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

axis(side = 3, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

# KBS0802
plot(KBS0802$time, KBS0802$Abund, pch = 22, 
          bg = "white", col = "black", lwd = 6, cex = 6, log = "y", xlab = NA, ylab = NA,
          xlim = c(-100, 1050), ylim = c(1, 1000000000), yaxt = "n", xaxt = "n")
box(lwd = 6)

axis(side = 2, lwd.ticks = 6, cex.axis = 3, las = 1,
     labels = expression(10^0, 10^2, 10^4, 10^6, 10^8),
     at = c(10^0, 10^2, 10^4, 10^6, 10^8))
 
axis(side = 4, lwd.ticks = 6, cex.axis = 3, las = 1,
    at=c(10^0, 10^2, 10^4, 10^6, 10^8), labels = F)
 
axis(side = 1, lwd.ticks = 6, cex.axis = 3, las = 1, mgp = c(3, 3, 0),
    labels = c(0, 500, 1000), at = c(0, 500, 1000))

axis(side = 3, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

# KBS0812
plot(KBS0812$time, KBS0812$Abund, pch = 22, 
          bg = "white", col = "black", lwd = 6, cex = 6, log = "y", xlab = NA, ylab = NA,
          xlim = c(-100, 1050), ylim = c(1, 1000000000), yaxt = "n", xaxt = "n")
box(lwd = 6)

axis(side = 2, lwd.ticks = 6, cex.axis = 3, las = 1,
 at=c(10^0, 10^2, 10^4, 10^6, 10^8), labels = F)
 
axis(side = 4, lwd.ticks = 6, cex.axis = 3, las = 1,
    at=c(10^0, 10^2, 10^4, 10^6, 10^8), labels = F)
 
axis(side = 1, lwd.ticks = 6, cex.axis = 3, las = 1, mgp = c(3, 3, 0),
    labels = c(0, 500, 1000), at = c(0, 500, 1000))

axis(side = 3, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

dev.off()
graphics.off()
```

## 5) WEIBULL FUNCTIONS
```{r, results='hide', warning=FALSE, message=FALSE}

## KBS0701
# Initial parameters
A = 1000 # Initial death (larger = slower) 
B = 0.9 # Bend (upper = 1 = first-order decay)
C = 6 # intercept
Z = 2 # Error

fit.0701 <- mle2(log10(Abund) ~ dnorm(mean = c * (time / a)^(b-1) * exp(-1*(time/a)^b), sd = z), 
            start = list(a = A, b = B, c = C, z = Z), data = KBS0701) 

## KBS0714
# Initial parameters
A = 100 # Initial death (larger = slower) 
B = 1 # Bend (upper = 1 = first-order decay)
C = 7 # intercept
Z = 2 # Error

fit.0714 <- mle2(log10(Abund) ~ dnorm(mean = c * (time / a)^(b-1) * exp(-1*(time/a)^b), sd = z), 
            start = list(a = A, b = B, c = C, z = Z), data = KBS0714) 

## KBS0802
# Initial parameters
A = 400 # Initial death (larger = slower) 
B = 0.8 # Bend (upper = 1 = first-order decay)
C = 4 # intercept
Z = 10 # Error

fit.0802 <- mle2(log10(Abund) ~ dnorm(mean = c * (time / a)^(b-1) * exp(-1*(time/a)^b), sd = z), 
            start = list(a = A, b = B, c = C, z = Z), data = KBS0802) 

## KBS0812
# Initial parameters
A = 9100 # Initial death (larger = slower) 3000
B = 1.1 # Bend (upper = 1 = first-order decay) 1.1
C = 7 # intercept 8 
Z = 100 # Error 10
fit.0812 <- mle2(log10(Abund) ~ dnorm(mean = c * (time / a)^(b-1) * exp(-1*(time/a)^b), sd = z), 
            start = list(a = A, b = B, c = C, z = Z), data = KBS0812) 

```

## 6) FOUR-PANEL PLOT WITHOUT FUNCTIONS
```{r}
png(filename="~/GitHub/Dimensions/Aim1/output/death.weibull.fitting.png",
    width = 2400, height = 2400, res = 96*2)

par(mfrow = c(2,2),
          oma = c(5,4,0,0) + 0.5,
          mar = c(0,3,3,1) + 0.5)

# KBS0701
plot(KBS0701$time, log10(KBS0701$Abund), pch = 22, 
          bg = "white", col = "black", lwd = 6, cex = 6, xlab = NA, ylab = NA,
          xlim = c(-100, 1050), ylim = c(0, 8.5), yaxt = "n", xaxt = "n")
box(lwd = 6)

axis(side = 2, lwd.ticks = 6, cex.axis = 3, las = 1,
     labels = expression(10^0, 10^2, 10^4, 10^6, 10^8),
     at = c(0, 2, 4, 6, 8))
 
axis(side = 4, lwd.ticks = 6, cex.axis = 3, las = 1,
    at=c(0, 2, 4, 6, 8), labels = F)
 
axis(side = 1, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

axis(side = 3, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

curve(coef(fit.0701)[[3]] * (x / coef(fit.0701)[[1]])^(coef(fit.0701)[[2]]-1) * exp(-1*(x/coef(fit.0701)[[1]])^coef(fit.0701)[[2]]), 
      from = 1, to = 1000, add = TRUE, lwd = 10, lty = 1, col = "red") 

# KBS0714
plot(KBS0714$time, log10(KBS0714$Abund), pch = 22, 
          bg = "white", col = "black", lwd = 6, cex = 6, xlab = NA, ylab = NA,
          xlim = c(-100, 1050), ylim = c(0, 8.5), yaxt = "n", xaxt = "n")
box(lwd = 6)

axis(side = 2, lwd.ticks = 6, cex.axis = 3, las = 1,
     at = c(0, 2, 4, 6, 8), labels = F)
 
axis(side = 4, lwd.ticks = 6, cex.axis = 3, las = 1,
    at=c(0, 2, 4, 6, 8), labels = F)
 
axis(side = 1, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

axis(side = 3, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

curve(coef(fit.0714)[[3]] * (x / coef(fit.0714)[[1]])^(coef(fit.0714)[[2]]-1) * exp(-1*(x/coef(fit.0714)[[1]])^coef(fit.0714)[[2]]), 
      from = 10, to = 275, add = TRUE, lwd = 10, lty = 1, col = "red") 

# KBS0802
plot(KBS0802$time, log10(KBS0802$Abund), pch = 22, 
          bg = "white", col = "black", lwd = 6, cex = 6, xlab = NA, ylab = NA,
          xlim = c(-100, 1050), ylim = c(0, 8.5), yaxt = "n", xaxt = "n")
box(lwd = 6)

axis(side = 2, lwd.ticks = 6, cex.axis = 3, las = 1,
     labels = expression(10^0, 10^2, 10^4, 10^6, 10^8),
     at = c(0, 2, 4, 6, 8))
 
axis(side = 4, lwd.ticks = 6, cex.axis = 3, las = 1,
    at=c(0, 2, 4, 6, 8), labels = F)
 
axis(side = 1, lwd.ticks = 6, cex.axis = 3, las = 1, mgp = c(3, 3, 0),
    labels = c(0, 500, 1000), at = c(0, 500, 1000))

axis(side = 3, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

curve(coef(fit.0802)[[3]] * (x / coef(fit.0802)[[1]])^(coef(fit.0802)[[2]]-1) * exp(-1*(x/coef(fit.0802)[[1]])^coef(fit.0802)[[2]]), 
      from = 1, to = 1000, add = TRUE, lwd = 10, lty = 1, col = "red") 

# KBS0812
plot(KBS0812$time, log10(KBS0812$Abund), pch = 22, 
          bg = "white", col = "black", lwd = 6, cex = 6, xlab = NA, ylab = NA,
          xlim = c(-100, 1050), ylim = c(0, 8.5), yaxt = "n", xaxt = "n")
box(lwd = 6)

axis(side = 2, lwd.ticks = 6, cex.axis = 3, las = 1,
     at = c(0, 2, 4, 6, 8), labels = F)
 
axis(side = 4, lwd.ticks = 6, cex.axis = 3, las = 1,
    at=c(0, 2, 4, 6, 8), labels = F)
 
axis(side = 1, lwd.ticks = 6, cex.axis = 3, las = 1, mgp = c(3, 2, 0),
    labels = c(0, 500, 1000), at = c(0, 500, 1000))

axis(side = 3, lwd.ticks = 6, cex.axis = 3, las = 1,
    at = c(0, 500, 1000), labels = F)

curve(coef(fit.0812)[[3]] * (x / coef(fit.0812)[[1]])^(coef(fit.0812)[[2]]-1) * exp(-1*(x/coef(fit.0812)[[1]])^coef(fit.0812)[[2]]), 
      from = 1, to = 1000, add = TRUE, lwd = 10, lty = 1, col = "red") 

dev.off()
graphics.off()
```