---
title: "Long-Term Dormancy"
author: "Stuart E. Jones and Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW
In January 2013, we started an experiment using ~24 strains of bacteria -- some "lab" strains others enviornmental strains -- to estimate survivorship during long-term starvation. Each strain was grown un into logarithmic phase and harvested. The cells were pelleted and washed 5x before being put in replicate 50 mL Falcon tubes. Subsamples were taken on overtime and plated onto R2A plates for enumeration as CFUs. After looking at plots of the data, in became apparent that CFUs were declining at a non-constant rate on semi-log plots. We therefore developed a statistical model to test for whether the the data were better fit by a first-order decay (i.e., log-linear) or a saturating function.  

## 1) SET WORKING DIRECTORY AND LOAD DATA
```{r}
## Retrieve and Set Your Working Directory
rm(list = ls())
getwd()
setwd("~/GitHub/Dimensions/Aim1/DeathCurves/")

## Load Data
obs <- read.csv("data/longtermdormancy_20151112_nocomments.csv", 
                header = TRUE, stringsAsFactors = FALSE)
```

## 2) ESTIMATE COLONY FORMING UNITS (CFU)
```{r}
## Adding 1 to deal with log(0) observations
obs$Abund <- as.numeric(obs$Colonies) * 10 ^ as.numeric(obs$Dilution) + 1
```

## 3) LIKELIHOOD FUNCTION FOR LINEAR FIT TO LOG-TRANSFORMED DATA
```{r}
fitLogLinearDecay<-function(p,N,time){
	k=p[1]
	N0=p[2]
	sd=exp(p[3])
	
	Nt=N0-k*time
	
	-sum(dnorm(N,Nt,sd,log=TRUE))
}
```

## 4) LIKELIHOOD FUNCTION FOR QUADRATIC FIT TO LOG-TRANSFORMED DATA
```{r}
fitLogQuadDecay<-function(p,N,time){
	k2=p[1]
	k=p[2]
	N0=p[3]
	sd=exp(p[4])

	Nt=N0-k2*time^2-k*time
	
	-sum(dnorm(N,Nt,sd,log=TRUE))
}
```

## 5) SELECT STRAINS WITH > 10 OBSERVATIONS 
```{r}
strains <- sort(unique(obs$Strain)) 
strains <- strains[table(obs$Strain)>10]
obs <- obs[obs$Strain%in%strains,]
```

## 6) FIT MODELS
```{r}
# Uncomment following line to create pdf that will plot data and fits
# pdf('decayFits.pdf') 

# Create matrix for storing model fit output
summ <- matrix(NA, length(strains) * max(obs$Rep), 10)

# Start model fitting
counter = 1

for(i in 1:length(strains)){
	strainObs=obs[obs$Strain==strains[i],]
	
	reps=unique(strainObs$Rep)
	for(j in 1:length(reps)){
		
		repObs=strainObs[strainObs$Rep==reps[j],]

		if(nrow(repObs)>3){
			
			start=repObs[1,1]
		
			time=(as.numeric(strptime(repObs$Firstread_date,format="%d-%b-%y",tz="EST"))-as.numeric(strptime(start,format="%d-%b-%y",tz="EST")))/(3600*24)
		# fits the log-linear model by maximum likelihood
			logLinCur=optim(c((log10(repObs$Abund[1])-log10(repObs$Abund[nrow(repObs)]))/(time[length(time)]-time[1]),log10(max(repObs$Abund)),1),fitLogLinearDecay,N=log10(repObs$Abund),time=time)
		# fits the log-quadratic model by maximum likelihood	
    tempQuad=optim(c(((log10(repObs$Abund[1])-log10(repObs$Abund[nrow(repObs)]))/(time[length(time)]-time[1]))/10,(log10(repObs$Abund[1])-log10(repObs$Abund[nrow(repObs)]))/(time[length(time)]-time[1]),log10(max(repObs$Abund)),1),fitLogQuadDecay,N=log10(repObs$Abund),time=time)
		# fits the log-quadratic model by maximum likelihood, but with different parms	
    tempQuad2=optim(c(-5e-6,0.005,5,log(0.1)),fitLogQuadDecay,N=log10(repObs$Abund),time=time)
		
			if(tempQuad2$value<=tempQuad$value){
				logQuadCur=tempQuad2
			}else{
				logQuadCur=tempQuad	
			}
		
			summ[counter,1]=strains[i]
			summ[counter,2]=reps[j]
			summ[counter,3:4]=logLinCur$par[1:2]
			summ[counter,5]=round(2*logLinCur$value+2*length(logLinCur$par),2)
			summ[counter,6:8]=logQuadCur$par[1:3]
			summ[counter,9]=round(2*logQuadCur$value+2*length(logQuadCur$par),2)
		
			# run likelihood ratio test to compare the linear and quadratic model; store p-value in summ
			summ[counter,10]=pchisq((2*logLinCur$value-2*logQuadCur$value),df=1,lower.tail=FALSE)
		
			#### add indicator of non-linearity (****) to plot title if quadratic model is better
#			if(!is.na(summ[counter,10])){
#				if(as.numeric(summ[counter,10])<0.05){
#					title=paste(strains[i]," rep ",reps[j],"****")
#				}else{
#					title=paste(strains[i]," rep ",reps[j])
#				}
#			}else{
#				title=paste(strains[i]," rep ",reps[j])
#			}
			
#			plot(time,log10(repObs$Abund),main=title,ylim=c(0,9))
#			predTime=seq(0,max(time))
#			lines(predTime,logLinCur$par[2]-logLinCur$par[1]*predTime,lwd=2,lty=2)
#			lines(predTime,logQuadCur$par[3]-logQuadCur$par[1]*predTime^2-logQuadCur$par[2]*predTime,col='red',lwd=2,lty=2)
		
			counter=counter+1
		}
	}
}

#dev.off()

summ=summ[!is.na(summ[,1]),]
colnames(summ)=c('strain','rep','linearfit_K','linearfit_N0','linearfit_AIC','quadfit_k2','quadfit_k','quadfit_N0','quadfit_AIC','LRT_pvalue')

#percent of tubes that are better fit by quadratic (AIC_linear > AIC_quadratic)
sum(as.numeric(summ[,5])>as.numeric(summ[,9]))/nrow(summ)*100	# 51.3%

# percent of tubes that are better based on likelihood ratio test
sum(as.numeric(summ[,10])<0.05)/nrow(summ)*100	# 46%

# add bonferroni correction
0.05/nrow(summ)	#  0.00044
sum(as.numeric(summ[,10])<0.0044)/nrow(summ)*100	# 31.9%

## histogram of likelihood ratio p-values
hist(log10(as.numeric(summ[,10])),breaks=seq(-20,0,1))
abline(v=log10(0.05),lwd=2,lty=2,col='red')
abline(v=log10(0.00044),lwd=2,lty=2,col='green')
legend('topleft',c('alpha=0.05','alpha=0.00044 (bonferroni)'),lty=2,col=c('red','green'),box.lty=0,lwd=2)

### confirm that likelihood ratio test is showing what deltaAIC between linear and quadratic would suggest
plot(log10(as.numeric(summ[,10])),as.numeric(summ[,9])-as.numeric(summ[,5]),xlab="log10(LRT pvalue)",ylab="deltaAIC (quadratic - linear)")
abline(v=log10(0.05),lwd=2,lty=2,col='red')
abline(v=log10(0.00044),lwd=2,lty=2,col='green')

### look at agreement amongst reps
repAgreement=matrix(NA,length(strains),4)
repAgreement[,1]=strains
for(i in 1:length(strains)){
	cur=summ[summ[,1]==strains[i],]
	
	repAgreement[i,2]=nrow(cur)
	repAgreement[i,3]=sum(as.numeric(cur[,10])>0.05)
	repAgreement[i,4]=nrow(cur)-sum(as.numeric(cur[,10])>0.05)
}

repAgree=data.frame(strain=repAgreement[,1],Nreps=as.numeric(repAgreement[,2]),Nlinear=as.numeric(repAgreement[,3]),Nquad=as.numeric(repAgreement[,4]),stringsAsFactors=FALSE)

### generate trait output table
# when quadratic is not significantly better by likelihood ratio test report "evolvability" of 0
traitsReport=matrix(NA,nrow(summ),5)
traitsReport[,1]=summ[,1]
traitsReport[,2]=summ[,2]
for(i in 1:nrow(summ)){
	if(as.numeric(summ[i,10])<(0.05/nrow(summ))){
		traitsReport[i,3]=summ[i,7]
		traitsReport[i,4]=summ[i,6]
		traitsReport[i,5]=summ[i,8]
	}else{
		traitsReport[i,3]=summ[i,3]
		traitsReport[i,4]=0
		traitsReport[i,5]=summ[i,4]
	}
}
colnames(traitsReport)=c('strain','rep','decay','evolvability','N0')

# remove the extremely bad fit
traitsReport=traitsReport[-27,]

strains=sort(unique(traitsReport[,1]))
meanDecay=tapply(as.numeric(traitsReport[,3]),traitsReport[,1],FUN=mean)
meanEvolve=tapply(as.numeric(traitsReport[,4]),traitsReport[,1],FUN=mean)
meanN0=tapply(as.numeric(traitsReport[,5]),traitsReport[,1],FUN=mean)

summaryTraitsReport=data.frame(strains=strains,decay=meanDecay,evolvability=meanEvolve,N0=meanN0,stringsAsFactors=FALSE)

write.table(traitsReport,"perRepDeathCurveTraits.txt",row.names=FALSE,quote=FALSE)
write.table(summaryTraitsReport,"DeathCurveTraits.txt",row.names=FALSE,quote=FALSE)
