---
title: "Long-term dormancy - proportion active"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

Estimating distribution of activity among samples at end of experiment

```{r}
rm(list=ls())
getwd()
setwd("/Users/lennonj/GitHub/Dimensions/Aim1/DeathCurves")
```

```{r}
require("png")
```

## Load Data

```{r}
stain.der <- read.table("data/staining.derived.txt", head = TRUE, sep = "\t")
stain.anc <- read.table("data/staining.ancestor.txt", head = TRUE, sep = "\t")
stain.all <- read.table("data/staining.all.txt", head = TRUE, sep = "\t")
```

## Take time differences
```{r}

strain.anc <- stain.all[which(stain.all$hist == "anc"), ]
strain.der <- stain.all[which(stain.all$hist == "der"), ]

try <- stain.all[which(stain.all$hist == "der" & stain.all$strain =="KBS0701"), 4] - stain.all[which(stain.all$hist == "anc" & stain.all$strain=="KBS0701"), 4]

for (row in 1:nrow(df)) { 
    df$newvar[row] <- testFunc(df$x[row], df$z[row]) 
}


try<- apply(stain.all[which(stain.all$hist == "der"), 4], 2, function (x) sum(x))



rich.rep.ave <- ddply(stain.all, .(strain, treat, qPCR.match), 
                      summarize, rich = mean(rich.mean))

rich.rep.ave <- ddply(eDNA.div, .(strain), 
                      summarize, diff = (rich.mean))

stain.all$hist == anc - stain.all$     by strain
## Make Kernels
```{r}
kern.der.act <- density(stain.der$Active, bw = 0.1) 
kern.der.dead <- density(stain.der$Dead, bw = 0.1)
kern.der.dorm <- density(stain.der$Dormant, bw = 0.1)
kern.anc.act <- density(stain.anc$Active, bw = 0.1) 
kern.anc.dead <- density(stain.anc$Dead, bw = 0.1) 
kern.anc.dorm <- density(stain.anc$Dormant, bw = 0.1) 
```

## Make Figure for Derived Srtains
```{r}
# Set-up for printing 
png(filename="/Users/lennonj/GitHub/Dimensions/Aim1/DeathCurves/figures/stain.derived.kernel.png",width = 1200, height = 900, res = 96*2)

# Set plotting parameters
plot.new()
par(mar = c(7, 7, 5, 7))
layout <- layout(cbind(1, 2, 3), height = c(2, 2, 2)) 
plot(kern.der.act, main = NA, xaxt = "n", yaxt = "n", cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-0.1, 1.1), ylim = c(0, 4.5), type = "l", lwd = 3, col = "red")

points(kern.der.dead, type = "l", lwd = 3, col = "darkgreen")
points(kern.der.dorm, type = "l", lwd = 3, col = "blue")

mtext('Density', side = 2, outer = TRUE, cex = 2.25, 
      line = -3, adj = 0.55)

mtext(expression('Metabolic Status (%)'), side = 1, outer = TRUE, 
      cex = 2.25, line = -3, adj = 0.5)
     
    axis(side = 1, lwd.ticks = 2, cex.axis = 2, las = 1,
        labels = c("0", "50", "100"), at = c(0, 0.5, 1.0))

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0), labels = F)

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 2, 4), labels = T)

    axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 2, 4), labels = F)

box(lwd = 2.25)

dev.off() # this writes plot to folder
graphics.off() # shuts down open devices
```

## Make 3-panel Figure for Ancestor vs. Derived Srtains
```{r}
# Set-up for printing 
png(filename="/Users/lennonj/GitHub/Dimensions/Aim1/DeathCurves/figures/stain.ancestor.derived.kernel.png", width = 1200, height = 600, res = 96*2)

layout <- layout(cbind(1, 2, 3), height = c(1, 1, 1)) 
#layout.show(layout)

# Plotting Active
#par(mar = c(7, 4, 2, 0.5), oma = c(1,1,1,1))
par(mar = c(6, 0.5, 5, 0.5))
par(oma = c(1, 5, 1, 2))
plot(kern.anc.act, xaxt = "n", yaxt = "n", main = NA, cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-0.1, 1.20), ylim = c(0, 4.5), type = "l", 
     lty = 1, lwd = 3, col = "red")

points(kern.der.act, type = "l", lty = 2, lwd = 3, col = "red")

     axis(side = 1, lwd.ticks = 2, cex.axis = 2, las = 1,
        labels = c("0", "50", "100"), at = c(0, 0.5, 1.0))

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0), labels = F)

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 2, 4), labels = F)

    # axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
     #   at = c(0, 2, 4), labels = F)

box(lwd = 2.25)

mtext("Density", side = 2, line = 2, cex = 2)
mtext("Active", side = 3, line = 1, cex = 1.25, col = "red")

legend("topright", c("ancestor", "derived"), bty = "n", y.intersp = 1, lty = c(1,2), 
       lwd = 2.5, seg.len = 2.75, col = "red", cex = 1)

# Add Dead

plot(kern.anc.dead, main = NA, xaxt = "n", yaxt = "n", cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-0.1, 1.20), ylim = c(0, 4.5), type = "l", 
     lty = 1, lwd = 3, col = "darkgreen")

points(kern.der.dead, type = "l", lty = 2, lwd = 3, col = "darkgreen")

axis(side = 1, lwd.ticks = 2, cex.axis = 2, las = 1,
        labels = c("0", "50", "100"), at = c(0, 0.5, 1.0))

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0), labels = F)

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 2, 4), labels = F)

  #  axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
   #     at = c(0, 2, 4), labels = F)

box(lwd = 2.25)

mtext("Dead", side = 3, line = 1, cex = 1.25, col = "darkgreen")
mtext("Individuals (%)", side = 1, line = 5, cex = 2)

legend("topright", c("ancestor", "derived"), bty = "n", y.intersp = 1, lty = c(1,2), 
       lwd = 2.5, seg.len = 2.5, col = "darkgreen", cex = 1)

# Add Dormant

plot(kern.anc.dorm, main = NA, xaxt = "n", yaxt = "n", cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-0.1, 1.20), ylim = c(0, 4.5), type = "l", 
     lty = 1, lwd = 3, col = "blue")

points(kern.der.dorm, type = "l", lty = 2, lwd = 3, col = "blue")

axis(side = 1, lwd.ticks = 2, cex.axis = 2, las = 1,
        labels = c("0", "50", "100"), at = c(0, 0.5, 1.0))

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0), labels = F)

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 2, 4), labels = F)

    axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 2, 4), labels = F)

box(lwd = 2.25)

mtext("Dormant", side = 3, line = 1, cex = 1.25, col = "blue")

legend("topright", c("ancestor", "derived"), bty = "n", y.intersp = 1, lty = c(1,2), 
       lwd = 2.5, seg.len = 2.75, col = "blue", cex = 1)

dev.off() # this writes plot to folder
graphics.off() # shuts down open devices
```
