---
title: "Long-term Dormancy Experiment - Phylogeny"
author: "Jay T. Lennon, Stuart E. Jones, and William R. Shoemaker"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## 1) SETUP
### A. Retrieve and Set Your Working Directory

```{r, results = 'hide'}
rm(list = ls())
getwd()
setwd("~/GitHub/LTDE/phylo")
```


### B. Load Packages 

```{r, results = 'hide', message = FALSE, warning = FALSE} 
require("seqinr")
require("ape")
require("phylobase")
require("adephylo")
require("geiger") 
require("picante")
require("RColorBrewer")
require("caper")
```


## 2) Use mothur to align the sequences using `silva.v4.fasta` as a reference. 

The code below is run on Mason as the shell script `persistence_707_712_721.sh`
```{bash, eval = FALSE}
#!/bin/bash
#PBS -k o
#PBS -l nodes=1:ppn=8,vmem=100gb,walltime=0:30:00
#PBS -M wrshoema@indiana.edu
#PBS -m abe
#PBS -j oe

module load gcc/4.9.2
module load boost/1.52.0
module load openmpi
module load mothur/1.38.1

cd /N/dc2/projects/Lennon_Sequences/LTDE_Tree

mothur persistence_707_712_721.batch
```

that calls the `persistence_707_712_721.batch` file, which contains the following lines
```{bash, eval = FALSE}
align.seqs(fasta=persistence_707_712_721.fasta, reference=silva.v4.fasta, flip=T, processors=4)
summary.seqs(fasta=persistence_707_712_721.align)
screen.seqs(fasta=persistence_707_712_721.align, minlength=200)
filter.seqs(fasta=persistence_707_712_721.good.align, vertical=T, trump=.)
```



## 3) Visualize alignments
```{r}
# Read mothur alignment file {seqinr}
read.aln.M <- read.alignment(file = "./align/persistence_707_712_721.good.filter.fasta", format = 'fasta')
# Convert Alignment File to DNAbin Object {ape}
p.DNAbin.M <- as.DNAbin(read.aln.M) 

# Identify Base Pair Region of 16S rRNA Gene to Visuzlize (adjust range)
window.M <- p.DNAbin.M

# Command to Visusalize Sequence Alignment {ape}
image.DNAbin(window.M, cex.lab = 0.50) 
```



## 4) Generate a ML tree with bootstrap support using RAxML
```{bash, eval = FALSE}
raxmlHPC-PTHREADS -T 4 -f a -m GTRGAMMA -p 12345 -x 12345 -o Methanosarcina -# autoMRE \
    -s /Users/WRShoemaker/GitHub/LTDE/phylo/tree/persistence_707_712_721.good.filter.fasta \
    -n T20 -w /Users/WRShoemaker/GitHub/LTDE/phylo/tree
```

## 5) Read in maximum likelihood tree
```{r}
# Read tree
ml.tree <- read.tree("./tree/RAxML_bipartitionsBranchLabels.T20")

# Identify Outgroup Sequence
outgroup <- match("Methanosarcina", ml.tree$tip.label)

# Root the Tree {ape}
ml.rooted <- root(ml.tree, outgroup, resolve.root = TRUE)

# Load phylo taxonomy data
tax <- read.table("persistence.phylo.txt", sep = "\t", header = TRUE)
rownames(tax) <- tax$Code
tax.2 <- tax[ml.tree$tip.label, ]
tax.name <- paste(tax.2$Code, tax.2$Genus)
tax.name[15] <- "DSM2834 Methanosarcina"

# Plot the Rooted Tree{ape}
par(mar = c(1,1,2,1) + 0.1)
plot.phylo(ml.rooted, main = "ML Fast Tree", 
           "phylogram", use.edge.length = FALSE, direction = "right",
           cex = 0.6, label.offset = 1, show.tip.label = FALSE, x.lim = 30)
           
tiplabels(tax.name, adj = c(0,0.5), cex = 0.5, frame = "none",
          pch = NULL, thermo = NULL, pie = NULL, piecol = NULL,
          col = NULL, bg = NULL)

nodelabels(ml.rooted$node.label, font = 2, bg = "white", frame = "r", cex = 0.5)
add.scale.bar(cex = 0.7)
```

