---
title: "Long-term dormancy: bet-hedging"
author: "Stuart E. Jones and Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin = 2.54cm
---

## Overview
We measured the single cell activity on cells from different populations under conditions with freshly supplied media after 1 vs. 1000 days. 
Single cell staining classified cells as either active (+ CTC), dead (+ SYTOX Green), or dormant (- CTC and - SYTOX Green).

## 1) SETUP

### A. Retrieve and Set Working Directory

```{r, results = 'hide'}
rm(list = ls())
getwd()
setwd("~/GitHub/Dimensions/Aim1/")
require(ks)
```

## 2) LOAD AND WRANLE DATA

```{r, results = 'hide'}
all <- read.table("data/staining.all.txt", header = TRUE, sep = "\t", 
    stringsAsFactors = FALSE)

# Parse ancestral and derived measures
anc <- all[all$hist == "anc",]
der <- all[all$hist == "der",]
```

Create new traits based on single-cell metabolic counts at the two time points
```{r}
# Create dataframe to hold new traits
newTraits = der

# Percent of non-CTC stained cells at day 1
newTraits$betHedge = NA

# Difference in CTC cells between day 1 and 1000
newTraits$deltaActive = NA

# Difference in Sytox Green cells between day 1 and 1000
newTraits$deltaDead = NA

# Difference in Sytox Green cells between day 1 and 1000
newTraits$deltaDormant = NA

newTraits = newTraits[,-(3:6)]

# Calculate traits --> NOTE: no ancestral measures for KBS0727

for(i in 1:nrow(der)){
	acur = anc[anc$strain == der$strain[i],]
	newTraits[i,3] = acur[1,6]
	newTraits[i,4:6] = der[i,4:6] - acur[1,4:6]
}

newTraits <- na.omit(newTraits)
```

## 3) STATISTICS

```{r}
# Are delta activities different from zero?
t.test.bet <- t.test(newTraits$betHedge, mu=0)
t.test.act <- t.test (newTraits$deltaActive, mu=0)
t.test.dead <- t.test (newTraits$deltaDead, mu=0)
t.test.dorm <- t.test (newTraits$deltaDormant, mu=0)
```

## 4) PLOT KERNEL DENSTIES

```{r, results = 'hide'}
# Proportion of dormant cells at time zero
h_hedge <- hns(newTraits$betHedge,deriv.order=1)
bet.hedge.kern <- density(newTraits$betHedge, h_hedge)
mean(newTraits$betHedge)

h_act <- hns(newTraits$deltaActive,deriv.order=1)
delta.act.kern <- density(newTraits$deltaActive, bw = h_act)

h_dead <- hns(newTraits$deltaDead,deriv.order=1)
delta.dead.kern <- density(newTraits$deltaDead, bw = h_dead)

h_dorm <- hns(newTraits$deltaActive,deriv.order=1)
delta.dorm.kern <- density(newTraits$deltaDormant, bw = h_dorm)

# https://stats.stackexchange.com/questions/33918/is-there-an-optimal-bandwidth-for-a-kernel-density-estimator-of-derivatives

# Used site above to estimate h, which was used as bandwidth for kernels

png(filename="~/GitHub/Dimensions/Aim1/output/KernelShift.png",
    width = 600, height = 1200, res = 96*2)

layout <- layout(rbind(1, 2, 3)) 

# par(mai = c(0.5, 1.0, 0.5, 1.0))

# Plotting Delta Active
par(mar = c(2,7,4,4))
plot(delta.act.kern, xaxt = "n", yaxt = "n", main = NA, cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 1.5), type = "l", 
     lty = 1, lwd = 3, col = "black")

     axis(side = 1, lwd.ticks = 2, cex.axis = 1.5, las = 1,
        labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 1.5, las = 1,
          labels = c("0", "0.5", "1.0", "1.5"), at = c(0, 0.5, 1.0, 1.5))

     axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0, 1.5), labels = F)

box(lwd = 2.25)

mtext("Density", side = 2, line = 4, cex = 1.5)
mtext("Active", side = 3, line = 1, cex = 1.25, col = "black")
abline(v = 0, lty = 2, lwd = 2)

# Plotting Delta Dead
par(mar = c(2,7,4,4))
plot(delta.dead.kern, main = NA, xaxt = "n", yaxt = "n", cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 1.5), type = "l", 
     lty = 1, lwd = 3, col = "black")

     axis(side = 1, lwd.ticks = 2, cex.axis = 1.5, las = 1,
        labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 1.5, las = 1,
          labels = c("0", "0.5", "1.0", "1.5"), at = c(0, 0.5, 1.0, 1.5))

     axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0, 1.5), labels = F)

box(lwd = 2.25)

mtext("Density", side = 2, line = 4, cex = 1.5)
mtext("Dead", side = 3, line = 1, cex = 1.25, col = "black")
abline(v = 0, lty = 2, lwd = 2)

# Plotting Delta Dormant
par(mar = c(2,7,4,4))
plot(delta.dorm.kern, main = NA, xaxt = "n", yaxt = "n", cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 1.5), type = "l", 
     lty = 1, lwd = 3, col = "black")

     axis(side = 1, lwd.ticks = 2, cex.axis = 1.5, las = 1,
        labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 1.5, las = 1,
          labels = c("0", "0.5", "1.0", "1.5"), at = c(0, 0.5, 1.0, 1.5))

     axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0, 1.5), labels = F)

box(lwd = 2.25)

mtext("Density", side = 2, line = 4, cex = 1.5)
mtext("Dormant", side = 3, line = 1, cex = 1.25, col = "black")
abline(v = 0, lty = 2, lwd = 2)

dev.off() # this writes plot to folder
graphics.off() # shuts down open devices
```

## 5) CALCULATE STRAIN MEANS
```{r, results = 'hide'}
#
meanTraits <- newTraits[1:length(unique(der$strain)),]
meanTraits <- meanTraits[,-2]
meanTraits$strain = sort(unique(der$strain))
meanTraits <- meanTraits[which(meanTraits$strain !="KBS0727" & meanTraits$strain !="KBS0725"),]

for(i in 2:5){
	curMeans <- tapply(newTraits[,(i+1)], newTraits$strain, FUN = mean)
	meanTraits[,i] = curMeans[order(names(curMeans))]
}
write.table(meanTraits, "~/GitHub/Dimensions/Aim1/data/bet.hedge.table.txt", 
            sep = "\t") 
```


## 6) PLOT KERNEL DENSTIES (INCLUDING BET-HEDGING) FOR TALKS

```{r, results = 'hide'}

png(filename="~/GitHub/Dimensions/Aim1/output/KernelShift-4.png",
    width = 900, height = 900, res = 96*2)

# Plotting Delta Active

par(mfrow = c(2,2),
          oma = c(5,4,0,0) + 0.5,
          mar = c(0,0,3,1) + 0.5)

plot(delta.act.kern, xaxt = "n", yaxt = "n", main = NA, cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 2.0), type = "l", 
     lty = 1, lwd = 3, col = "black")

     # axis(side = 1, lwd.ticks = 2, cex.axis = 1.5, las = 1,
     #    labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 1, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 1.5, las = 1,
          labels = c("0", "1", "2"), at = c(0, 1.0, 2.0))

     axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 1.0, 2.0), labels = F)

box(lwd = 2.25)

#mtext("Density", side = 2, line = 4, cex = 1.5, adj = 0.5)
mtext("Active", side = 3, line = 1, cex = 1.25, col = "black")
abline(v = 0, lty = 2, lwd = 2)
mtext("Density", side = 2, cex = 2, line = 3, at = -0.5)

# Plotting Delta Dead

plot(delta.dead.kern, main = NA, xaxt = "n", yaxt = "n", cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 2), type = "l", 
     lty = 1, lwd = 3, col = "black")

     # axis(side = 1, lwd.ticks = 2, cex.axis = 1.5, las = 1,
     #    labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 1, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     # axis(side = 2, lwd.ticks = 2.25, cex.axis = 1.5, las = 1,
     #      labels = c("0", "1", "2"), at = c(0, 1.0, 2.0))

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 1.0, 2.0), labels = F)
     
     axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 1.0, 2.0), labels = F)

box(lwd = 2.25)

#mtext("Density", side = 2, line = 4, cex = 1.5)
mtext("Dead", side = 3, line = 1, cex = 1.25, col = "black")
abline(v = 0, lty = 2, lwd = 2)

# Plotting Delta Dormant

plot(delta.dorm.kern, main = NA, xaxt = "n", yaxt = "n", cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 2), type = "l", 
     lty = 1, lwd = 3, col = "black")

     axis(side = 1, lwd.ticks = 2, cex.axis = 1.5, las = 1,
        labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 2, lwd.ticks = 2.25, cex.axis = 1.5, las = 1,
          labels = c("0","1", "2"), at = c(0, 1.0, 2.0))

     axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 1.0, 2.0), labels = F)

box(lwd = 2.25)

#mtext("Density", side = 2, line = 4, cex = 1.5)
mtext("Dormant", side = 3, line = 1, cex = 1.25, col = "black")
abline(v = 0, lty = 2, lwd = 2)

# Plotting Bet-Hedging

plot(bet.hedge.kern, main = NA, xaxt = "n", yaxt = "n", cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 2.0), type = "l", 
     lty = 1, lwd = 3, col = "black")

     axis(side = 1, lwd.ticks = 2, cex.axis = 1.5, las = 1,
        labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 3, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)
     
     axis(side = 2, lwd.ticks = 2.25, cex.axis = 1.5, las = 1,
          at = c(0, 1.0, 2.0), labels = F)

     # axis(side = 2, lwd.ticks = 2.25, cex.axis = 1.5, las = 1,
     #      labels = c("0","1", "2"), at = c(0, 1.0, 2.0))

     axis(side = 4, lwd.ticks = 2.25, cex.axis = 2, las = 1,
        at = c(0, 1.0, 2.0), labels = F)

box(lwd = 2.25)

# mtext("Density", side = 2, line = 4, cex = 1.5)
mtext("Bet-hedging", side = 3, line = 1, cex = 1.25, col = "black")
abline(v = 0, lty = 2, lwd = 2)



dev.off() # this writes plot to folder
graphics.off() # shuts down open devices
```

## 7) PLOT ROW-WISE
```{r}

png(filename="~/GitHub/Dimensions/Aim1/output/KernelShift.row.png",
    width = 1200, height = 600, res = 96*2)
plot.new()
layout <- layout(cbind(1, 2, 3), height = c(1, 1, 1)) 
par(mar = c(6, 0.5, 5, 0.5) + 0.5)
par(oma = c(1, 5, 1, 2) + 0.5)

# par(mfrow = c(2,2),
#           oma = c(5,4,0,0) + 0.5,
#           mar = c(0,0,3,1) + 0.5)

# ACTIVE
plot(delta.act.kern, xaxt = "n", yaxt = "n", main = NA, cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 1.5), type = "l", 
     lty = 1, lwd = 3.5, col = "red")

     axis(side = 1, lwd.ticks = 3.5, cex.axis = 1.5, las = 1,
        labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 3, lwd.ticks = 3.5, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 2, lwd.ticks = 3.5, cex.axis = 1.5, las = 1,
          labels = c("0.0", "0.5", "1.0", "1.5"), at = c(0, 0.5, 1.0, 1.5))

     axis(side = 4, lwd.ticks = 3.5, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0, 1.5), labels = F)

mtext("Active", side = 3, line = 1, cex = 1.75, col = "black")
abline(v = 0, lty = 2, lwd = 3.5)
mtext("Density", side = 2, cex = 2, line = 4, at = 0.75)
     
box(lwd = 3.5)


# DEAD 
plot(delta.dead.kern, xaxt = "n", yaxt = "n", main = NA, cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 1.5), type = "l", 
     lty = 1, lwd = 3.5, col = "chartreuse3")

     axis(side = 1, lwd.ticks = 3.5, cex.axis = 1.5, las = 1,
        labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 3, lwd.ticks = 3.5, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 2, lwd.ticks = 3.5, cex.axis = 1.5, las = 1,
           at = c(0, 0.5, 1.0, 1.5), labels = F)

     axis(side = 4, lwd.ticks = 3.5, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0, 1.5), labels = F)

mtext("Dead", side = 3, line = 1, cex = 1.75, col = "black")
abline(v = 0, lty = 2, lwd = 3.5)

     
box(lwd = 3.5)

# DORMANT 
plot(delta.dorm.kern, xaxt = "n", yaxt = "n", main = NA, cex.lab = 1.5, ylab = "",
     xlab = "", xlim = c(-1.5, 1.5), ylim = c(0, 1.5), type = "l", 
     lty = 1, lwd = 3.5, col = "blue")

     axis(side = 1, lwd.ticks = 3.5, cex.axis = 1.5, las = 1,
        labels = c("-1", "0", "1"), at = c(-1, 0, 1))

     axis(side = 3, lwd.ticks = 3.5, cex.axis = 2, las = 1,
        at = c(-1, 0, 1), labels = F)

     axis(side = 2, lwd.ticks = 3.5, cex.axis = 1.5, las = 1,
           at = c(0, 0.5, 1.0, 1.5), labels = F)

     axis(side = 4, lwd.ticks = 3.5, cex.axis = 2, las = 1,
        at = c(0, 0.5, 1.0, 1.5), labels = F)

mtext("Dormant", side = 3, line = 1, cex = 1.75, col = "black")
abline(v = 0, lty = 2, lwd = 3.5)

     
box(lwd = 3.5)

dev.off() # this writes plot to folder
graphics.off() # shuts down open devices

```
